Glyph-maps for Visually Exploring Temporal Patterns in Climate Data and Models

### 2012-03-04 revision/paper.tex
abstract and intro

@@

Understanding the multiple aspects of climate data is a challenge. There are both spatial and temporal components to examine, and often multivariate dependencies. This paper proposes a new display for climate data: glyph-maps. Glyph-maps are a specialization of multivariate glyph plots; each spatial location is displayed with one glyph that represents the measurements recorded over time at that location. Glyph-maps allow the discovery of both local and global structure, with a particular focus on temporal relationships.

The instructions for computing and displaying glyph-maps are described, along with the design of reference frames, and the effects of different scaling of the data on perception of structure. The methods are developed for rectangularly gridded data, but with some clever processing, which is explained, it is possible get good glyph-maps of irregularly gridded data.

@@

Climate data is composed of measurements, such as temperature, precipitation, and winds, with a spatial and temporal context. The classic display for data of this type of is ``heatmap'', a tiled plot using color to display the value of the variable of interest at each location. When measurements are made at multiple time points it is common to display the data with small multiples, such as in Figure~\ref{fig:nasa-facet}. In this plot, a separate map is drawn for each month (columns) and year (rows), over a six year period of remotely sensed temperature data above Central America \citep{murrell:2010}. Color is used to display de-seasonalized temperature, with red mapped to high values and blue to low. The most noticeable feature is the strong red patch in the equatorial Pacific beginning mid-1997 and tapering out during 1998. This is the El Ni\~no event, a major temperature anomaly. More work is required to see more localized patterns, like the cooler land temperatures seen in early years.

Reading this type of plot is cognitively challenging: the reader must play ``spot the difference'' from one small image to another. Large structures such as El Ni\~no are clear but it is very difficult to mentally difference the images and read off the long-term trend or to notice local deviations. Using a movie to render these small multiples can help, but small trends still fail to draw readers' attention and escape unnoticed \citep{simons:gradual}.

@@

This paper develops a new type of display to resolve these problems: the \emph{glyph-map}. The glyph-map uses a small glyph, or icon, to represent multiple values at each location. Glyph maps are an adaptation of glyphs, tools developed for display multivariate data. Multivariate glyphs include the star plots of \citet{mayr:1877}, the semi-graphic displays of \citet{anderson:1960}, and the infamous Chernoff faces \citep{chernoff:1973}. An glyph is produced by mapping each variable to some graphical feature, such as the length of a line. The glyph-map can also be thought of as a small multiple display \citep{tufte:2001} of time series with a geographic context.

Figure~\ref{fig:nasa-glyph} shows the glyph-map equivalent of Figure~\ref{fig:nasa-facet}: each location is represented by a small time series glyph. The primary visible structure is still El Ni\~no, the regions in the Pacific with bump in the middle of the series, but other features are also visible: temperatures are generally more varied over land and a few locations have dramatically increasing temperature in the later years typically in high elevation areas, along the Andes and the Sierra Madre Occidental in Mexico.

>>>>

% resolution issues with large data - reference - aloso long time series.

% emphasize way of thinking, which leads into interactive setting

@@

The multivariate spatio-temporal nature of climate data makes it difficult to draw all of the aspects simultaneously. This paper describes the conceptualization and construction of a type of display, a glyph-map, that has been used infrequently for climate data. Glyph-maps are a specialization of multivariate glyph plots. Each spatial location is displayed with one glyph that represents the multiple measurements, recorded over time, at that location. Glyph-maps allow the discovery of both local and global structure, with a particular focus on temporal relationships. They provide alternatives to colored, facetted maps, or statistical summaries such as principal components. With the new conceptualization, developing interactive versions of glyph-maps will be easier, and it will be simpler to explore the perceptual effects of different scales. The methods are developed for rectangularly gridded data, but with some clever processing, which is explained, it is possible get good glyph-maps of irregularly gridded data.

%Understanding the multiple aspects of climate data is a challenge. There are both spatial and temporal components to examine, and often multivariate dependencies. This paper proposes a new display for climate data: glyph-maps. Glyph-maps are a specialization of multivariate glyph plots; each spatial location is displayed with one glyph that represents the measurements recorded over time at that location. Glyph-maps allow the discovery of both local and global structure, with a particular focus on temporal relationships.

%The instructions for computing and displaying glyph-maps are described, along with the design of reference frames, and the effects of different scaling of the data on perception of structure. The methods are developed for rectangularly gridded data, but with some clever processing, which is explained, it is possible get good glyph-maps of irregularly gridded data.

@@

Climate data is composed of multiple variables, such as temperature, ozone, precipitation, and winds, with a spatial and temporal context. The classic display for data of this type of is ``heatmap'', a tiled plot using color to display the value of the variable of interest at each location. When measurements are made at multiple time points it is reasonably common to display the data with small multiples~\citep{tufte:2001,IPCC}, such as in Figure~\ref{fig:nasa-facet}. In this plot, a separate map is drawn for each month (columns) and year (rows), over a six year period of remotely sensed temperature data above Central America \citep{murrell:2010}. Color is used to display de-seasonalized temperature, with red mapped to high values and blue to low. The most noticeable feature is the strong red patch in the equatorial Pacific beginning mid-1997 and tapering out during 1998. This is the El Ni\~no event, a major temperature anomaly (see for example, \citet{trenberth} or \citet{ElNino2}.) More work is required to see more localized patterns, like the cooler land temperatures seen in early years.

To read long-term or localized temporal patterns, this type of plot is cognitively challenging: the reader must play ``spot the difference'' from one small image to another. Large spatial structures such as El Ni\~no are clear but it is very difficult to extract the temporal pattern because it requires mentally differencing the images. Using a movie to render these small multiples can help, but small trends still fail to draw readers' attention and escape unnoticed \citep{simons:gradual}. The \emph{glyph-map} display described in this paper enables easier perception of temporal patterns.

@@

The glyph-map uses a small glyph, or icon, to represent multiple variables, or time values, at each location. Glyph maps derive from glyph plots, which were initially developed to display multivariate data. One glyph is drawn for each observation in the data set. Types of glyph plots include the star plots \citep{mayr:1877, kleiner:1981}, the semi-graphic displays \citep{anderson:1960}, and the infamous Chernoff faces \citep{chernoff:1973}. An glyph is produced by mapping each variable to some graphical feature, such as the length of a line, curvature of the mouth, or size of the face.

Figure~\ref{fig:nasa-glyph} shows the glyph-map equivalent of Figure~\ref{fig:nasa-facet}: each location is represented by a small time series glyph of de-seasonalized temperature. The darker grey in the background represents the land mass of the study region. Because the data is de-seasonalized, the primary visible structure is still El Ni\~no -- it is both a temporal anomaly and a spatial phenomenon. It can be recognized by the bump in the middle of the series of the equatorial Pacific region. Other temporal features are visible from the many time series. Temperatures are generally more varied over land, than sea. A few locations have dramatic jumps in temperature in the later years, all in high elevation areas, along the Andes and the Sierra Madre Occidental in Mexico.

The glyph-map can also be thought of as a small multiple display \citep{tufte:2001}. The primary comparison is different. Cognitively, elements placed close to each other are more easily and accurately compared \citep{cleveland:1984}. The facetted colored map, places greater emphasis on comparison between the spatial locations, because these are positioned together, and secondary emphasis on time because these are separated by facets. In the glyph-map primary emphasis is placed on the temporal comparison, and secondarily on spatial comparison, because facetting is on spatial locations.
